Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-10 paper/section-instantiations.tex
Inhabitation of AG

\exref{ex:type-iso} is a special case of this isomorphism; another instance is the
type of vector addition:
\begin{eqnarray*}
\forall t_1\mathord:\SynTransl{2}.\forall t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_1+t_2}\\
\cong\quad
\forall t_1\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_1}\quad\cong\quad
\tyPrim{vec}{0}
\end{eqnarray*}
\noindent
\newcommand{\transup}[1]{\uparrow_{#1}}
\newcommand{\transdn}[1]{\downarrow_{#1}}
The maps associated with this isomorphism are 
$I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transdn A (y)$,
%\begin{eqnarray*}
%I = \lambda x. x\ [0]\ 0 & \text{ and } &
%J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
%\end{eqnarray*}
where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A$ have definition symmetric to $\transdn A$):
\begin{eqnarray*}
\transdn{1} & = & \lambda x. x\\ 
\transdn{A_1\times A_2} & = & \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
\transdn{A_1 + A_2} & = & \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y)); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
\transdn{A_1\to A_2} & = & \lambda f. \lambda x. (\transdn{A_2} (f(\transup{A_1}(x))))\\ 
\transdn{\tyPrim{vec}{e}} & = & \lambda x. x - kv\\ 
\transup{\tyPrim{vec}{e}} & = & \lambda x. x + kv\\ & & \text{ if }e \equiv k t + e'\text{ for
  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'\\
\end{eqnarray*}
To show $I(J(y))$ is contextually equivalent to $y$ is an easy
induction on $A$ using the fact that the vector $0$
is an identity for vector addition. To show $J(I(x))$ is contextually
equivalent to $x$ we must apply the abstraction theorem and use
the fact that $\tmSem{\transup A}$ and $\tmSem{\transdn A}$ 
considered as relations are simply instances of $\rsem A{}{}$.
%\[
%$\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\transup A (x)}) \mid x \in
%\tySem A\}$.
%\]

@@

\subsection{Indefinability}

\fixme{Copy the stuff over from the other file}

>>>>

\exref{ex:type-iso} is a special case of this isomorphism; another instance is the
type of vector addition:
\begin{eqnarray*}
\forall t_1\mathord:\SynTransl{2}.\forall t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_1+t_2}\\
\cong\quad
\forall t_1\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_1}\quad\cong\quad
\tyPrim{vec}{0}
\end{eqnarray*}
\noindent
\newcommand{\transup}[1]{\uparrow_{#1}}
\newcommand{\transdn}[1]{\downarrow_{#1}}
The maps associated with this isomorphism are 
$I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transup A (y)$,
%\begin{eqnarray*}
%I = \lambda x. x\ [0]\ 0 & \text{ and } &
%J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
%\end{eqnarray*}
where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A$ have definition symmetric to $\transdn A$):
\begin{eqnarray*}
\transdn{1} & = & \lambda x. x\\ 
\transdn{A_1\times A_2} & = & \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
\transdn{A_1 + A_2} & = & \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y)); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
\transdn{A_1\to A_2} & = & \lambda f. \lambda x. (\transdn{A_2} (f(\transup{A_1}(x))))\\ 
\transdn{\tyPrim{vec}{e}} & = & \lambda x. x - kv\\ 
\transup{\tyPrim{vec}{e}} & = & \lambda x. x + kv\\ & & \text{ if }e \equiv k t + e'\text{ for
  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'\\
\end{eqnarray*}
To show $I(J(y))$ is contextually equivalent to $y$ is an easy
induction on $A$ using the fact that the vector $0$
is an identity for vector addition. To show $J(I(x))$ is contextually
equivalent to $x$ we must apply the abstraction theorem and use
the fact that $\tmSem{\transup A}$ and $\tmSem{\transdn A}$ 
considered as relations are simply instances of $\rsem A{}{}$.
%\[
%$\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\transup A (x)}) \mid x \in
%\tySem A\}$.
%\]

@@

\subsection{Indefinability}
\label{sec:types-indexed-abelian-groups-indef}
\newcommand{\Grp}{\mathit{Grp}}
For proving non-definability results, we
consider the general case of a sort $\mathsf{G}$ with the equational theory of abelian groups.
We assume a group-indexed primitive type $\tyPrimNm{G}$ and
set of primitive operations typed by $\Gamma_\Grp$ that reflect the group structure at the value level:
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
    1 & \tyPrim{G}{1} \\
    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a\cdot b} \\
    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
  \end{array}
\end{displaymath}
One instance is the sort $\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ operations.

For index-erasure interpretation, we assume a (not necessarily abelian) group $(G, 1, \cdot,
\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
elements that commute with every other element in $G$. We set
$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
interpreted just as the unit, multiplication and inverse operations
of $G$. For each index context $\Delta$, the set
$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
\to Z(G) \}$, where
\begin{displaymath}
  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
    \begin{array}{ll}
      \{ \} & \textrm{if}\ e\not\in H \\
      \{ (x, h(e)\oplus x) \sepbar x \in G \} & \textrm{if}\ e\in H
    \end{array}
  \right.
\end{displaymath}

\begin{lemma}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Grp(\Delta)$,
  $(\eta_\Grp,\eta_\Grp) \in \rsem{\Gamma_\Grp}{\relEnv{E}}\rho$.
\end{lemma}

We write $i^n$ for $n$-fold iteration of the group operation.
\begin{theorem}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
  there exists a program $M$ such that
  \begin{equation}
    \label{eq:fo-group-type}
    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
  \end{equation}
  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, if and only if there is a solution $\vec{x}$
  in integers to the equation $A \vec x = b$.
\end{theorem}
The uninhabited type  $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$
from \exref{ex:uninhabited-type} is an instance of this theorem.

For types indexed by the scaling group $\SynGL{1}$ the situation is a little more complex, as
the index-erasure interpretation of $\tyPrim{real}{s}$ is larger than the group $\GL{1}$, as it contains
zero. We can show, nonetheless, that $\forall s \mathord:
  \SynGL{1}.\ \tyPrim{real}{s \circ s} \to \tyPrim{real}{s}\cong 1$, reflecting 
the fact that any term of this type must be the function $\lambda x.0$.

%\begin{proof}
%  (If) The program $\Lambda i_1\dots i_m.\ \lambda g_1\dots g_n.\
%  g_1^{x_1}\dots g_n^{x_n}$ satisfies the typing judgment
%  (\ref{eq:fo-group-type}).

%  (Only if) Assume that there is a program $M$ satisfying the typing
%  judgment (\ref{eq:fo-group-type}). By \thmref{thm:abstraction} we
%  know that for all relational environments $\rho \in
%  \relEnv{E}_\Grp(i_1,\dots,i_m)$, and all $g_j,g'_j \in G$,
%  \begin{displaymath}
%    \begin{array}{l}
%      (g_1,g'_1) \in \rho\ \tyPrimNm{G}\ (e_1) \land \dots \land (g_n,g'_n) \in \rho\ \tyPrimNm{G}\ (e_n) \Rightarrow \\
%      \quad (\tmSem{M}g_1\dots g_n, \tmSem{M}g'_1\dots g'_n) \in \rho\ \tyPrimNm{G}\ (e)
%    \end{array}
%  \end{displaymath}
%  We select the relational environments
%  \begin{displaymath}
%    \rho\ \tyPrimNm{G}\ (e) = \left\{
%      \begin{array}{ll}
%        \{(g,g) \sepbar g \in \mathcal{G} \} &
%        \begin{array}[t]{@{}l}
%          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
%          \textrm{ for some }x_1,\dots,x_n \in \mathbb{Z}
%        \end{array}
%        \\
%        \{\} & \textrm{otherwise}
%      \end{array}
%    \right.
%  \end{displaymath}
%  Now, for any $g_1,\dots,g_m \in \mathcal{G}$, we know that, for all
%  $j$, $(g_j,g_j) \in \rho\ \tyPrimNm{G}\ (e_j)$, so we know that
%  $(\tySem{M}g_1\dots g_n, \tySem{M}g_1 \dots g_n) \in \rho\
%  \tyPrimNm{G}\ (e)$. Thus there exist integers $x_1,...,x_n$ such
%  that $e = e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
%  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free groups,
%  we learn that $x_1,...,x_n$ is the solution we need.
%\end{proof}
